Effectful Semantics for Substructural SSA

### 2023-11-29 pldi24/paper.tex
Tap at refinement

@@

In general, we often want to justify optimizations which \textit{refine} a
program by, for example, reducing the set of allowable behaviours, rather than
simply rewrite the program to an equivalent one. One place where refinement
pops up quite often is in the treatment of undefined values and undefined 
behaviour; for example, in a language where reading a dangling pointer is UB,

\TODO{freeze example, UB}

\TODO{determinism and local refinement}

@@

\TODO{semantics of refinement via poset-enriched categories}

\TODO{trivial poset-enrichment}

\TODO{powerset-based poset-enrichment}

>>>>

\label{ssec:substruct}

@@

In general, we often want to justify optimizations which \textit{refine} a
program by, for example, reducing the set of allowable behaviours, rather than
simply rewrite the program to an equivalent one. One place where refinement
pops up quite often is in the treatment of undefined values and undefined 
behaviour; for example, in a language where reading a dangling pointer is UB, 
the transformation 
\begin{equation*}
\begin{BVerbatim}[baseline=c, fontsize=\small]
  let x = read(p);
  ret 5
\end{BVerbatim}
  \qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c, fontsize=\small]
  ret 5
\end{BVerbatim}
\end{equation*}
is valid, but the reverse is \textit{not}, since that would \textit{introduce},
rather than \textit{remove}, potential UB. These optimizations are particularly
important when operations are allowed to generate undefined values, in which
case, even simple rewrites such as \(a \times 0 \to 0\) and \(a - a \to 0\) are
refinements rather than equalities. In this appendix section, we will generalize
our semantics for rewriting from section \ref{apx:rewriting} to support
reasoning about refinements in a compositional manner.

@@

To reason about refinement, we need to introduce the definition of a
\textit{poset-enriched category}:
\begin{definition}[Poset-Enriched Category]
  A category \(\mc{C}\) is \textbf{poset-enriched} if, for all objects \(A, B
  \in |\mc{C}|\), \(\mc{C}(A, B)\) is equipped with a partial order \(\leq\)
  compatible with composition, i.e. \(f \leq g \implies f;h \leq g;h, h';f \leq
  h';g\). A \textbf{poset-enriched premonoidal category} is a poset-enriched
  category such that the partial order on morphisms is compatible with
  whiskering, i.e. \(f \leq g \implies f \otimes C \leq g \otimes C, C \otimes f
  \leq C \otimes g\). A \textbf{poset-enriched substructural category} is a
  poset-enriched premonoidal category such that the subcategories
  \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\) are downward-closed under \(\leq\). 
\end{definition}
Note that every (premonoidal/substructural) category \(\mc{C}\) can be
considered poset-enriched under the discrete partial order \((\leq) = (=)\). One
useful example of a poset-enriched substructural category is the category of
relations from section \ref{ssec:substruct}, taking \(f \leq g \iff \forall a
\in A, f(a) \subseteq g(a)\).
